---
title: "Getting Started with rlichess"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rlichess}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`rlichess` is a modern, tidy R client designed for chess players and data scientists to interact with the [Lichess API](https://lichess.org/api) and analyze chess data using standard `tidyverse` principles.

## Package Architecture

`rlichess` is structured into three clean layers:

1. **Layer 1: API Clients (`lic_*`)**: Direct, transparent HTTP wrappers corresponding 1-to-1 with official Lichess API endpoints.
2. **Layer 2: Tidy Wranglers (`lic_tidy_*`)**: Pure offline transformation functions that clean, unnest, and standardize chess data into tidy tibbles.
3. **Layer 3: Chess Analytics (`lic_stats_*`)**: Offline statistics and analysis tools (e.g., opening win rates).

```
[ Lichess REST API ]
        │
        ▼ (Layer 1: API Clients)
  lic_games_user() / lic_user() / lic_explorer_lichess()
        │
        ▼ (Layer 2: Tidy Wranglers)
  lic_tidy_games() / lic_tidy_moves()
        │
        ▼ (Layer 3: Analytics / Visualization)
  lic_stats_openings() / dplyr / ggplot2
```

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("h8gi/rlichess")

# or using devtools
devtools::install_github("h8gi/rlichess")
```

```{r setup}
library(rlichess)
library(dplyr)
```

## 1. Authentication (Optional but Recommended)

While public endpoints can be accessed anonymously, providing a Personal Access Token increases rate limits (up to 30–60 games/sec) and enables on-demand rating history retrieval.

Set your token in your `.Renviron` file:

```bash
LICHESS_API_TOKEN="your_personal_access_token"
```

In R code, `rlichess` automatically loads it via `lic_token()`:

```{r}
lic_token()
```

## 2. Layer 1: API Clients

### Fetching User Games

`lic_games_user()` downloads game records directly from the Lichess API in NDJSON format:

```{r}
# Download recent bullet games since 2025
raw_games <- lic_games_user(
  username = "h8gi",
  perf_type = "bullet",
  since = "2025-01-01",
  max = 50,
  clocks = TRUE,
  evals = TRUE
)
```

### Fetching User Profiles & Rating History

```{r}
# User profile summary as a 1-row tidy tibble
user <- lic_user("h8gi")
user %>% select(username, created_at, play_time_total_hours, count_all)

# Performance ratings across game types
perfs <- lic_user_perfs("h8gi")
perfs

# Daily rating history
history <- lic_user_rating_history("h8gi", perf_type = "bullet")
```

### Querying Opening Explorer Databases

```{r}
# Query Lichess opening database (e.g. 1. e4 c5)
exp <- lic_explorer_lichess(play = "e4,c5")
exp$moves

# Query FIDE Masters database (2200+ FIDE)
masters <- lic_explorer_masters(play = "e4,c5")
masters$moves

# Query a specific player's repertoire
player_exp <- lic_explorer_player("h8gi", color = "white", play = "e4")
player_exp$moves
```

## 3. Layer 2: Tidy Wranglers

### Standardizing Game Data

`lic_tidy_games()` parses raw API output into a user-centric format with `POSIXct` timestamps, user perspective (`user_color`, `user_result`, `win`), opponent information, and rating differences:

```{r}
games <- lic_tidy_games(raw_games, username = "h8gi")

games %>%
  select(id, created_at, user_color, user_result, user_rating, opponent_name, opening.name) %>%
  head()
```

### Expanding Move Sequences (Long Format)

For move-by-move analysis (evaluations, clock times, blunders), `lic_tidy_moves()` unnests games into a long format where each row is a single half-move (ply):

```{r}
moves_df <- lic_tidy_moves(games)

moves_df %>%
  filter(game_id == first(game_id)) %>%
  select(ply, move_number, color, san, clock, eval) %>%
  head(10)
```

## 4. Layer 3: Chess Analytics

### Opening Win Rates

Calculate win rates and sample sizes for each opening variation:

```{r}
stats <- lic_stats_openings(games, min_games = 5)
stats
```

## 5. Built-in Dataset

Access the included database of 3,378 ECO openings:

```{r}
data(lichess_openings)
head(lichess_openings)
```
