---
title: "Get started with rlichess"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with rlichess}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

**rlichess** is a modern, tidy R client designed for chess players, analysts, and researchers. It bridges the [Lichess REST & Streaming API](https://lichess.org/api) with the `tidyverse` ecosystem, allowing you to seamlessly fetch games, user statistics, opening databases, and tactical puzzles into tidy tibbles.

---

## 1. Package Architecture

`rlichess` separates data access, data cleaning, and statistical calculations into three clear layers:

```
┌────────────────────────────────────────────────────────┐
│ Layer 1: API Clients (Network / HTTP Requests)          │
│   • lic_games_user()      • lic_game()                 │
│   • lic_user()            • lic_user_rating_history()  │
│   • lic_explorer_*()      • lic_puzzle_*()             │
└──────────────────────────┬─────────────────────────────┘
                           │ Raw tibble / list
                           ▼
┌────────────────────────────────────────────────────────┐
│ Layer 2: Tidy Wranglers (Offline Data Transformation)  │
│   • lic_tidy_games()  (user-centric perspective/dates) │
│   • lic_tidy_moves()  (1-row-per-ply move unnesting)   │
└──────────────────────────┬─────────────────────────────┘
                           │ Tidy tibble
                           ▼
┌────────────────────────────────────────────────────────┐
│ Layer 3: Chess Analytics & Visualization               │
│   • lic_stats_openings()                               │
│   • standard dplyr / tidyr / ggplot2 workflows         │
└────────────────────────────────────────────────────────┘
```

---

## 2. Installation & Setup

Install the latest version from GitHub:

```r
# install.packages("pak")
pak::pak("h8gi/rlichess")

# or using devtools / remotes
devtools::install_github("h8gi/rlichess")
```

```{r setup}
library(rlichess)
library(dplyr)
```

### Personal API Token (Optional)

While public endpoints can be accessed anonymously, configuring a Lichess Personal Access Token increases rate limits (up to 30–60 games/second) and enables on-demand rating history generation.

Add your token to your `.Renviron` file:

```bash
LICHESS_API_TOKEN="lip_your_personal_access_token_here"
```

In R code, `rlichess` automatically picks it up via `lic_token()`:

```{r}
lic_token()
```

---

## 3. Layer 1: API Clients

### Fetching User Games

`lic_games_user()` downloads game records directly from Lichess in NDJSON format. You can filter by date range, time control, opponent, or computer analysis:

```{r}
# Download recent bullet games since 2025
raw_games <- lic_games_user(
  username = "h8gi",
  perf_type = "bullet",
  since = "2025-01-01",
  max = 100,
  clocks = TRUE,
  evals = TRUE
)
```

### Downloading a Single Game or Multiple Games by IDs

```r
# Single game
game <- lic_game("0tMlsM69")

# Batch export multiple games by IDs (up to 300)
batch_games <- lic_games_export_ids(c("0tMlsM69", "q7ZvsdUF"))
```

### User Profiles, Status, Crosstable & Leaderboards

```r
# Core user profile summary as a 1-row tidy tibble
user <- lic_user("h8gi")
user %>%
  select(username, created_at, play_time_total_hours, count_all)

# Check real-time online and playing status for multiple users
statuses <- lic_users_status(c("h8gi", "magnuscarlsen", "hikaru"), with_game_ids = TRUE)
statuses

# Head-to-head match crosstable between two players
crosstable <- lic_user_crosstable("Lance5500", "TryingHard87")
crosstable

# Leaderboard / top players ranking
top_blitz <- lic_leaderboard(perf_type = "blitz", count = 10)
top_blitz

# Performance ratings across all game types (bullet, blitz, rapid, puzzle, etc.)
perfs <- lic_user_perfs("h8gi")
perfs

# Daily rating history (dates and rating values)
history <- lic_user_rating_history("h8gi", perf_type = "bullet")

# In-depth performance statistics (streaks, best wins, worst losses)
stats <- lic_user_perf_stats("h8gi", perf = "bullet")
```

### Study PGN Export

```r
# Export full study PGN (or specific chapter)
pgn <- lic_study_pgn("sK9kZ8vT")
```

### Opening Explorer Databases

Query Lichess's vast opening database (over billions of games) or historical FIDE master-level games (2200+ FIDE):

```r
# Query Sicilian Defense (1. e4 c5) in Lichess database
exp <- lic_explorer_lichess(play = "e4,c5")
exp$moves

# Query FIDE master tournament games
masters <- lic_explorer_masters(play = "e4,c5")
masters$moves

# Query a specific player's personal opening repertoire
player_exp <- lic_explorer_player("h8gi", color = "white", play = "e4")
player_exp$moves
```

### Daily Tactical Puzzle

```r
daily <- lic_puzzle_daily()
daily$puzzle$rating
```

---

## 4. Layer 2: Tidy Wranglers

Raw API data often contains nested JSON objects and unformatted move strings. Layer 2 provides offline transformation functions to clean and reshape this data.

### Tidying Game Metadata (`lic_tidy_games`)

`lic_tidy_games()` parses raw game data into a user-centric perspective:
- Converts timestamps into standard `POSIXct` datetime vectors (`created_at`, `last_move_at`).
- Identifies the player's color (`user_color`) and match result (`user_result`, `win`).
- Extracts opponent names (`opponent_name`), opponent ratings (`opponent_rating`), and rating differences (`user_rating_diff`).

```{r}
games <- lic_tidy_games(raw_games, username = "h8gi")

games %>%
  select(id, created_at, user_color, user_result, user_rating, opponent_name, opening.name) %>%
  head()
```

### Unpacking Move Sequences (`lic_tidy_moves`)

To analyze move-by-move evaluations, clock times, or time management, `lic_tidy_moves()` unpacks game move strings into a long-format tibble where **each row corresponds to one half-move (ply)**:

```{r}
moves_df <- lic_tidy_moves(games)

moves_df %>%
  filter(game_id == first(game_id)) %>%
  select(ply, move_number, color, san, clock, eval) %>%
  head(10)
```

---

## 5. Layer 3: Chess Analytics

### Opening Win Rates

Summarize performance across different opening variations:

```{r}
opening_stats <- lic_stats_openings(games, min_games = 5)
opening_stats
```

---

## 6. Built-in Dataset: `lichess_openings`

`rlichess` bundles the official Lichess opening database containing 3,378 ECO codes, opening names, and move sequences:

```{r}
data(lichess_openings)
head(lichess_openings)
```
